\thetitle

### 2025-05-13 workshop/PT.tex
compress

Pigeons.jl provides an implementation of distributed parallel tempering (PT) described 
in \cite{syed2021nrpt}, which we outline in Algorithm~\ref{alg:distributed_PT}.
This section gives both a brief overview of PT and some details of 
its distributed implementation.

@@

In this section we assume a basic understanding of 
Markov chain Monte Carlo (MCMC) methods.
For readers unfamiliar with MCMC, it is important to know that 
it is a method to obtain approximate samples from a distribution. 
However, with traditional 
MCMC methods, the samples may be heavily correlated instead of independent 
and may fail to sufficiently explore the 
full space of the distribution (see \cref{fig:bimodal}); 
PT is a method that aims to address these issues.
For a more in-depth review of PT, we refer readers to \cite{surjanovic2022vpt}.

\subsection{Overview of PT}
Suppose that we would like to estimate integrals involving $\pi$, such as 
$\int f(x) \pi(x) \, \dee x$. 
These integrals may be multivariate and even include combinations of 
continuous and discrete variables (where sums replace integrals in the discrete 
case).
One method is to obtain samples from $\pi$ to approximate such integrals.
Often, the distribution 
$\pi$ can be challenging for traditional MCMC methods---such 
as Metropolis-Hastings, slice sampling, and Hamiltonian Monte Carlo---
because of its structure.
For example, in a bimodal example such as the one illustrated in 
Figure~\ref{fig:bimodal},
traditional methods might remain in one of the two modes for an extremely 
long period of time.

To resolve this issue, PT constructs a sequence of $N$ distributions,  
$\pi_1, \pi_2, \ldots, \pi_N$, where $\pi_N$ is usually equal to $\pi$.
The distributions are chosen so that it is easy to obtain samples from $\pi_1$
with the sampling difficulty increasing as one approches $\pi_N$. An example 
of such a sequence of distributions, referred to as an \textit{annealing path},
is shown in Figure~\ref{fig:path}.

We now turn to explain how this path of distributions can be used to enhance 
sampling from the target distribution, $\pi$. 
PT operates by first obtaining samples from each distribution on the path in parallel 
(referred to as an \textit{exploration phase}). 
Then, samples between adjacent distributions are swapped (referred to as 
a \textit{communication phase}). The communication phase in PT is crucial: it allows 
for the discovery of new regions of the space of the target distribution such as 
the top-right mode of the distribution presented in Figure~\ref{fig:bimodal}.

@@

% NB: this is placed here just to make it appear in a different page
\begin{figure}[t]
    \centering
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_1.pdf}
      \caption*{$\pi_1$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_2.pdf}
      \caption*{$\pi_2$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_3.pdf}
      \caption*{$\pi_3$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_4.pdf}
      \caption*{$\pi_4$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_5.pdf}
      \caption*{$\pi_5$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_6.pdf}
      \caption*{$\pi_6$}
    \end{minipage}
    \caption{Heatmaps of six distributions lying on an annealing path 
    from a unimodal reference distribution  
    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
    $\pi_N$, which is in this case the bimodal distribution from 
    Figure~\ref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
    are not directly comparable because the densities of intermediate distributions 
    are not normalized.)}
    \label{fig:path}
\end{figure}

>>>>

Pigeons.jl provides an implementation of distributed parallel tempering (PT) described 
in \cite{syed2021nrpt}, which we outline in \cref{alg:distributed_PT}.
This section gives both a brief overview of PT and some details of 
its distributed implementation.

@@

In this section we assume a basic understanding of 
Markov chain Monte Carlo (MCMC) methods.
For readers unfamiliar with MCMC, it is important to know that 
it is a method to obtain approximate samples from a distribution. 
However, with traditional 
MCMC methods, the samples may be heavily correlated instead of independent 
and may fail to sufficiently explore the 
full space of the distribution (see \cref{fig:bimodal}); 
PT is a method that aims to address these issues.
For a more in-depth review of PT, we refer readers to \citet{surjanovic2022vpt}.

\subsection{Overview of PT}
Suppose that we would like to estimate integrals involving $\pi$, such as 
$\int f(x) \pi(x) \, \dee x$. 
These integrals may be multivariate and even include combinations of 
continuous and discrete variables (where sums replace integrals in the discrete 
case).
One method is to obtain samples from $\pi$ to approximate such integrals.
Often, the distribution 
$\pi$ can be challenging for traditional MCMC methods---such 
as Metropolis-Hastings, slice sampling, and Hamiltonian Monte Carlo---
because of its structure.
For example, in a bimodal example such as the one illustrated in 
\cref{fig:bimodal},
traditional methods might remain in one of the two modes for an extremely 
long period of time.

To resolve this issue, PT constructs a sequence of $N$ distributions,  
$\pi_1, \pi_2, \ldots, \pi_N$, where $\pi_N$ is usually equal to $\pi$.
The distributions are chosen so that it is easy to obtain samples from $\pi_1$
with the sampling difficulty increasing as one approches $\pi_N$. An example 
of such a sequence of distributions, referred to as an \textit{annealing path},
is shown in \cref{fig:path}.

We now turn to explain how this path of distributions can be used to enhance 
sampling from the target distribution, $\pi$. 
PT operates by first obtaining samples from each distribution on the path in parallel 
(referred to as an \textit{exploration phase}). 
Then, samples between adjacent distributions are swapped (referred to as 
a \textit{communication phase}). The communication phase in PT is crucial: it allows 
for the discovery of new regions of the space of the target distribution such as 
the top-right mode of the distribution presented in \cref{fig:bimodal}.

@@

% NB: this is placed here just to make it appear in a different page
\begin{figure}[t]
    \centering
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_1.pdf}
      \caption*{$\pi_1$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_2.pdf}
      \caption*{$\pi_2$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_3.pdf}
      \caption*{$\pi_3$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_4.pdf}
      \caption*{$\pi_4$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_5.pdf}
      \caption*{$\pi_5$}
    \end{minipage}
    \begin{minipage}{0.15\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/heatmap_path_6.pdf}
      \caption*{$\pi_6$}
    \end{minipage}
    \caption{Heatmaps of six distributions lying on an annealing path 
    from a unimodal reference distribution  
    $\pi_1$, from which it is straightforward to obtain samples, and ending at 
    $\pi_N$, which is in this case the bimodal distribution from 
    \cref{fig:bimodal}. (Note that in this case the colours between the heatmaps 
    are not directly comparable because the densities of intermediate distributions 
    are not normalized.)}
    \label{fig:path}
\end{figure}

### 2025-05-13 workshop/introduction.tex
compress

Pigeons.jl\footnote{The source code for Pigeons.jl can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl}. 
Pigeons.jl v0.2.0, used in the example scripts below, is compatible with Julia 1.8+.}
enables users to sample efficiently from challenging probability distributions 
and solve integration problems by 
implementing state-of-the-art sampling algorithms \cite{syed2021nrpt,surjanovic2022vpt} 
that leverage distributed computation. Its simple user interface facilitates performing such computations 
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1000 processes running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Importantly, Pigeons comes with guarantees on \emph{strong parallelism invariance} (SPI), 
wherein the output for a given seed is \emph{identical} irrespective of the number 
of threads or processes. Such a level of reproducibility is rare in distributed 
software but of great use for the purposes of debugging in the context of sampling 
algorithms, which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

\section{Overview of Pigeons.jl}
A typical instance of the class of problems that Pigeons.jl tackles can be described by
a probability density\footnote{This density may also be a probability mass function 
(discrete variables). We also allow for a combination of discrete/continuous variables.} 
$\pi(x)$ called the \emph{target}. In many problems, e.g. in Bayesian statistics, the 
density $\pi$ is typically known only up to a normalizing constant, 
\[
\label{eq:normalizing_constant}
  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
\]  
where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
Pigeons.jl takes as input the function $\gamma$.

>>>>

Pigeons.jl\footnote{\url{https://github.com/Julia-Tempering/Pigeons.jl}}
enables users to sample efficiently from challenging probability distributions.
To achieve this, Pigeons.jl provides a multi-threaded and distributed implementation
of non-reversible parallel tempering
(PT; \citealp{syed2021nrpt,syed2021paths,surjanovic2022vpt,surjanovic2024ergodicity}), 
a state-of-the-art MCMC algorithm (see \cref{sec:PT} for an outline of the method).
Crucially, Pigeons.jl follows the tuning guidelines described in \citet{syed2021nrpt} 
to provide a hyperparameter-free implementation.
Pigeons.jl's simple user interface facilitates running PT
single-threaded, multi-threaded, and/or distributed over MPI-communicating machines. 
We have stress-tested Pigeons successfully with up to 1000 processes running 
concurrently on a compute cluster (Sockeye) at the University of British Columbia.
Importantly, Pigeons comes with guarantees on \emph{strong parallelism invariance} (SPI), 
wherein the output for a given seed is \emph{identical} irrespective of the number 
of threads or processes. Such a level of reproducibility is rare in distributed 
software but of great use for the purposes of debugging in the context of sampling 
algorithms, which produce stochastic output.
Specifically, Pigeons.jl is designed to be suitable and yield reproducible output for:
\begin{enumerate}
    \item one machine running on one thread;
    \item one machine running on several threads;
    \item several machines running, each using one thread, and
    \item several machines running, each using several threads.
\end{enumerate}

\section{Overview of Pigeons.jl}
A typical instance of the class of problems that Pigeons.jl tackles can be described by
a probability density (or mass) function
%\footnote{This density may also be a probability mass function 
%(discrete variables). We also allow for a combination of discrete/continuous variables.} 
$\pi(x)$ called the \emph{target}. In many problems, e.g. in Bayesian statistics, the 
density $\pi$ is typically known only up to a normalizing constant, 
\[
\label{eq:normalizing_constant}
  \pi(x) = \frac{\gamma(x)}{Z}, \qquad Z = \int \gamma(x) \, \dee x,
\]  
where $\gamma$ can be evaluated pointwise but $Z$ is typically unknown.
Pigeons.jl takes as input the function $\gamma$.
